---
title: "Class05 Data Visualization with ggplot2"
author: "Lance"
format: pdf
---


##Using GGPLOT

The ggplot package needs to be installed as it does not come with R "out of the box"

We use the `install.packages()` function to do this.

To use ggplot I need to load it up before I can call any of the functions in the package.  I do this with the `library()` function.

```{r}
library(ggplot2)
ggplot(cars)
```

All ggplot figures have at least 3 things:
-data (the stuff we want to plot)
-aesthetic mapping (aes values)
-geoms

```{r}
ggplot(cars)+
  aes(x=speed, y=dist)
```




```{r}
ggplot(cars) +
  aes(x = speed, y=dist)+
  geom_point()+
  labs(title = "Speed and Stopping Distance of Cars",
       x = "Speed(MPH)", 
       y = "Stopping Distance (ft)",
       subtitle = "faster cars need more distance to stop",
       caption = "Dataset: 'cars'")+
  geom_smooth(method="lm", se = FALSE)
```


ggplot is not the only graphing system in R there are lots of others. There is even base R graphics.
```{r}
plot(cars)
```

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
nrow(genes)
ncol(genes)
table(genes$State)
sum(genes$State == "up")/(sum(genes$State == "down")+sum(genes$State == "unchanging")+ sum(genes$State == "up"))
```

```{r}
ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()
```

```{r}
p <- ggplot(genes)+
  aes(x=Condition1, y=Condition2, col=State)+
  geom_point()+ 
  scale_color_manual(values = c("blue","gray","red"))+
  labs(title = "Gene Expression Changes Upon Drug Treatment",
       x = "Control(No Drug)", 
       y = "Drug Treatment")
       
p
```

```{r}
library(gapminder)
library(dplyr)
gapminder_2007 <- gapminder %>% filter(year==2007)
ggplot(gapminder_2007)+
  aes(x = gdpPercap, y=lifeExp, color=continent, size=pop)+
  geom_point(alpha=0.5)
```

```{r}
ggplot (gapminder_2007)+
  aes(x=gdpPercap, y = lifeExp, color = pop)+
  geom_point(alpha=0.8)
```

```{r}
ggplot(gapminder_2007) + 
  aes(x = gdpPercap, y = lifeExp, size = pop) +
  geom_point(alpha=0.5)
```

```{r}
ggplot(gapminder_2007) + 
  geom_point(aes(x = gdpPercap, y = lifeExp,
                 size = pop), alpha=0.5) + 
  scale_size_area(max_size = 10)
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957)

ggplot(gapminder_1957)+
aes(x = gdpPercap, y = lifeExp,size = pop, color=continent)+
geom_point(alpha=0.7) + 
scale_size_area(max_size = 15)
 
 
```

```{r}
gapminder_1957 <- gapminder %>% filter(year==1957 | year == 2007)

ggplot(gapminder_1957)+
aes(x = gdpPercap, y = lifeExp,size = pop, color=continent)+
geom_point(alpha=0.7) + 
scale_size_area(max_size = 15)+
facet_wrap(~year)
```

```{r}
gapminder_top5 <- gapminder %>% 
  filter(year==2007) %>% 
  arrange(desc(pop)) %>% 
  top_n(5, pop)

gapminder_top5
```

```{r}
ggplot (gapminder_top5)+
  geom_col(aes(x= country, y = pop))
```

```{r}
ggplot (gapminder_top5)+
  geom_col(aes(x= country, y = lifeExp))
```

```{r}
ggplot(gapminder_top5)+
  geom_col(aes(x = country, y= pop, fill=continent))
```

```{r}
ggplot(gapminder_top5)+
  geom_col(aes(x = country, y = pop, fill=lifeExp))
```

```{r}
ggplot(gapminder_top5)+
  geom_col(aes(x = country, y = pop, fill=gdpPercap))
```

```{r}
ggplot(gapminder_top5)+
  geom_col(aes(x=reorder(country, -pop), y = pop, fill=gdpPercap))
```

```{r}
ggplot(gapminder_top5)+
  geom_col(aes(x=reorder(country, -pop), y = pop, fill=country, col="gray30")) +
  guides(fill="none")
```
What is the significance of aesthetic as a function of geometry vs. on it's own?

```{r}
ggplot(gapminder_top5)+
aes(x=reorder(country, -pop), y = pop, fill=country)+
geom_col(col="gray30") +
  guides(fill="none")
```

##Flipping bar charts

```{r}
USArrests$State <- rownames(USArrests)
ggplot(USArrests) +
  aes(x=reorder(State,Murder), y=Murder) +
  geom_col() +
  coord_flip()
```

```{r}
ggplot(USArrests)+
  aes(x=reorder(State,Murder), y=Murder)+
  geom_point() +
  geom_segment(aes(x=State,xend=State, y=0, yend=Murder), color="blue")+
  coord_flip()
  
```

##9. Extensions: Animation
```{r}
library(gapminder)
library(gganimate)

ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  shadow_wake(wake_length = 0.1, alpha = FALSE)
```

##10. Combining Plots
```{r}
library(patchwork)

p1 <- ggplot(mtcars) + geom_point(aes(mpg, disp))
p2 <- ggplot(mtcars) + geom_boxplot(aes(gear, disp, group = gear))
p3 <- ggplot(mtcars) + geom_smooth(aes(disp, qsec))
p4 <- ggplot(mtcars) + geom_bar(aes(carb))
(p1 | p2 | p3) /
      p4
```








